---
title: "Data comparability"
author: "Molly Wilson"
date: "3/26/2019"
output: html_document
---

```{r, echo=FALSE,message=FALSE,warning=FALSE}
library(tidyverse)
library(ggplot2)
library(here)
library(janitor)
```


```{r, echo=F, warning=F, message=F}
fish <- read.csv(here('data','raw_2019.csv'), stringsAsFactors = F) %>% 
  clean_names() %>%
  mutate(number = replace_na(number, 1)) %>% # change count to 1 if number was not entered
  uncount(number) %>% # expand to replicate rows if multiple fish were recorded
  mutate(year = as.factor(year)) %>%
  filter(code != "brc", code != "blc") %>% # removing chromis for consistency
  filter(site != "Barcadera", site != "Calabas", site != "Forest") # filter out sites that MWW did not survey, plus Calabas where BDR had only 1 transect

dodge <- position_dodge(width=0.9) # for bar graphs
```

```{r, echo=F, warning=F, message=F}
# total fish biomass, by site and surveyor
surv.tran <- fish %>% 
  group_by(surveyor, site, transect) %>% 
  summarize(bm_t = sum(biomass_fb)/1.2, 
            den_t = n()/1.2, 
            l_t = mean(length_cm))
surv.site <- surv.tran %>% 
  group_by(surveyor, site) %>% 
  summarize(n = n(), 
            bm_s = mean(bm_t), bm_se = (sd(bm_t))/sqrt(n), 
            den_s = mean(den_t), den_se = (sd(den_t))/sqrt(n), 
            l_s = mean(l_t), l_se = (sd(l_t))/sqrt(n)) # average across transects within each site

# herbivore biomass, by site and surveyor
herb_fam <- c("Scaridae","Acanthuridae") # families of interest
h.surv.tran <- fish %>% 
  filter(family %in% herb_fam) %>%
  group_by(surveyor, site, transect) %>% 
  summarize(bm_t = sum(biomass_bh)/1.2, # use Bohnsack & Harper biomass for consistency with past reports
            den_t = n()/1.2, 
            l_t = mean(length_cm))
h.surv.site <- h.surv.tran %>% 
  group_by(surveyor, site) %>% 
  summarize(n = n(), 
            bm_s = mean(bm_t), bm_se = (sd(bm_t))/sqrt(n), 
            den_s = mean(den_t), den_se = (sd(den_t))/sqrt(n), 
            l_s = mean(l_t), l_se = (sd(l_t))/sqrt(n)) # average across transects within each site

# carnivore biomass, by site and surveyor
carn_grp <- c("Carnivore","Omnivore")
c.surv.tran <- fish %>% 
  filter(functional_group %in% carn_grp) %>%
  group_by(surveyor, site, transect) %>% 
  summarize(bm_t = sum(biomass_fb)/1.2, # use Bohnsack & Harper biomass for consistency with past reports
            den_t = n()/1.2, 
            l_t = mean(length_cm))
c.surv.site <- c.surv.tran %>% 
  group_by(surveyor, site) %>% 
  summarize(n = n(), 
            bm_s = mean(bm_t), bm_se = (sd(bm_t))/sqrt(n), 
            den_s = mean(den_t), den_se = (sd(den_t))/sqrt(n), 
            l_s = mean(l_t), l_se = (sd(l_t))/sqrt(n)) # average across transects within each site

```

Notes: 

- BDR had a large school of creole wrasse (130 at 17cm each) and school of smallmouth grunts (33 at 17cm each) at Reef Scientifico (MWW had schools of both species but with fewer fish within the transect)  
- BDR had large school (36) of acanthurids and a large barracuda (64cm) at Karpata  
- BDR had school of black durgon (13 at 27cm each) at NDR, MWW had black durgon schools but mostly smaller individuals  

```{r, echo=F, warning=F, message=F}
ggplot(surv.site, aes(x = site, y = bm_s, group = surveyor, fill = surveyor)) +
 geom_bar(position = dodge, stat = "identity") + 
  geom_errorbar(aes(ymax = bm_s + bm_se, ymin = bm_s - bm_se), position = dodge, width = 0.25) +
  ggtitle("Total fish biomass") +
  ylab("Biomass (g/100m2)") +
  scale_fill_manual(values = c("slategray1","slategray")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r, echo=F, warning=F, message=F}
ggplot(surv.site, aes(x = site, y = bm_s, group = surveyor)) +
 geom_point(aes(color = factor(surveyor), shape = factor(surveyor))) + 
  geom_errorbar(aes(ymax = bm_s + bm_se, ymin = bm_s - bm_se, color = factor(surveyor)), width = 0.25) +
  ggtitle("Total fish biomass") +
  ylab("Biomass (g/100m2)") +
  scale_color_manual(values = c("aquamarine3","skyblue4")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r, echo=F, warning=F, message=F}
ggplot(surv.site %>%
         filter(site != "Barcadera", site != "Calabas", site != "Forest"),
       aes(x = surveyor, y = bm_s, group = site)) +
 geom_point(aes(color = factor(site), shape = factor(site))) + 
 geom_line(aes(color = factor(site))) +
  scale_shape_manual(values = c(0:8)) +
  # geom_errorbar(aes(ymax = bm_s + bm_se, ymin = bm_s - bm_se, color = factor(site)), width = 0.25) +
  ggtitle("Total fish biomass") +
  ylab("Biomass (g/100m2)") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r, echo=F, warning=F, message=F}
ggplot(surv.site, aes(x = site, y = den_s, group = surveyor, fill = surveyor)) +
 geom_bar(position = dodge, stat = "identity") + 
  geom_errorbar(aes(ymax = den_s + den_se, ymin = den_s - den_se), position = dodge, width = 0.25) +
  ggtitle("Total fish density") +
  ylab("Density (indv./100m2)") +
  scale_fill_manual(values = c("slategray1","slategray")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r, echo=F, warning=F, message=F}
ggplot(surv.site, aes(x = site, y = den_s, group = surveyor)) +
 geom_point(aes(color = factor(surveyor), shape = factor(surveyor))) + 
  geom_errorbar(aes(ymax = den_s + den_se, ymin = den_s - den_se, color = factor(surveyor)), width = 0.25) +
  ggtitle("Total fish density") +
  ylab("Density (indv./100m2)") +
  scale_color_manual(values = c("aquamarine3","skyblue4")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r, echo=F, warning=F, message=F}
ggplot(surv.site %>%
         filter(site != "Barcadera", site != "Calabas", site != "Forest"),
       aes(x = surveyor, y = den_s, group = site)) +
 geom_point(aes(color = factor(site), shape = factor(site))) + 
 geom_line(aes(color = factor(site))) +
  scale_shape_manual(values = c(0:8)) +
  ggtitle("Total fish density") +
  ylab("Density (indv./100m2)") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```



```{r, echo=F, warning=F, message=F}
ggplot(surv.site, aes(x = site, y = den_s, group = surveyor)) +
 geom_point(aes(color = factor(surveyor), shape = factor(surveyor))) + 
  geom_errorbar(aes(ymax = den_s + den_se, ymin = den_s - den_se, color = factor(surveyor)), width = 0.25) +
  ggtitle("Total fish density") +
  ylab("Density (indv./100m2)") +
  scale_color_manual(values = c("aquamarine3","skyblue4")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
dif.bm <- with(surv.site, 
        bm_s[surveyor == "MWW"] - bm_s[surveyor == "BDR"])
shapiro.test(dif.bm) # normal distribution

ttest.bm <- t.test(bm_s ~ surveyor, data = surv.site, paired = TRUE)
ttest.bm
```

```{r}
dif.den <- with(surv.site, 
        den_s[surveyor == "MWW"] - den_s[surveyor == "BDR"])
shapiro.test(dif.den) # normal distribution

ttest.den <- t.test(den_s ~ surveyor, data = surv.site, paired = TRUE)
ttest.den
```

```{r}
dif.len <- with(surv.site, 
        l_s[surveyor == "MWW"] - l_s[surveyor == "BDR"])
shapiro.test(dif.len) # normal distribution

ttest.len <- t.test(l_s ~ surveyor, data = surv.site, paired = TRUE)
ttest.len 
```
Paired t-test results: no significant difference between surveyors for total fish biomass, total fish density, nor average fish length (though nearly significant for biomass, with BDR having a higher mean than MWW)

### Herbivores

```{r, echo=F, warning=F, message=F}
ggplot(h.surv.site, aes(x = site, y = bm_s, group = surveyor, fill = surveyor)) +
 geom_bar(position = dodge, stat = "identity") + 
  geom_errorbar(aes(ymax = bm_s + bm_se, ymin = bm_s - bm_se), position = dodge, width = 0.25) +
  ggtitle("Herbivore biomass") +
  ylab("Biomass (g/100m2)") +
  scale_fill_manual(values = c("slategray1","slategray")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r, echo=F, warning=F, message=F}
ggplot(h.surv.site, aes(x = site, y = den_s, group = surveyor, fill = surveyor)) +
 geom_bar(position = dodge, stat = "identity") + 
  geom_errorbar(aes(ymax = den_s + den_se, ymin = den_s - den_se), position = dodge, width = 0.25) +
  ggtitle("Herbivore density") +
  ylab("Density (indv./100m2)") +
  scale_fill_manual(values = c("slategray1","slategray")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 90, hjust = 1))
```

### Carnivores + omnivores

```{r, echo=F, warning=F, message=F}
ggplot(c.surv.site, aes(x = site, y = bm_s, group = surveyor, fill = surveyor)) +
 geom_bar(position = dodge, stat = "identity") + 
  geom_errorbar(aes(ymax = bm_s + bm_se, ymin = bm_s - bm_se), position = dodge, width = 0.25) +
  ggtitle("Carnivore + Omnivore biomass") +
  ylab("Biomass (g/100m2)") +
  scale_fill_manual(values = c("slategray1","slategray")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r, echo=F, warning=F, message=F}
ggplot(c.surv.site, aes(x = site, y = den_s, group = surveyor, fill = surveyor)) +
 geom_bar(position = dodge, stat = "identity") + 
  geom_errorbar(aes(ymax = den_s + den_se, ymin = den_s - den_se), position = dodge, width = 0.25) +
  ggtitle("Carnivore + Omnivore density") +
  ylab("Density (indv./100m2)") +
  scale_fill_manual(values = c("slategray1","slategray")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 90, hjust = 1))
```